    \documentclass[12pt]{article}

    \usepackage{amsmath,amssymb}
    \usepackage[colorlinks=true]{hyperref}
    \usepackage[shortlabels]{enumitem}
    \usepackage{verbatim}

    \begin{document}

    \title{Exercises: inference in the linear Gaussian model}
    \author{Joaqu\'{i}n Rapela\thanks{j.rapela@ucl.ac.uk}}

    \maketitle

    \section{Inferring location of a static submarine from its sonar
    measurements}

    A static submarine is located in a 2D planar surfarce deep in the sea. A
    priori, we model its unknown location with a 2D Gaussian random variable
    $\mathbf{z}$ with mean $\mu_z$ and covariance $\Sigma_z$.

    We obtain noisy measurements of the location of the submarine with a sonar.
    We represent a 2D sonar measurement with random variable $\mathbf{y}_n$,
    with $p(\mathbf{y}_n|\mathbf{z})=\mathcal{N}(y_n|\mathbf{z}, \Sigma_y)$.

    \begin{enumerate}[(a)]

        \item Sample 100 a priory locations of the submarine
            (i.e., $\mathbf{z}_1,\ldots,\mathbf{z}_{100}$) using
            a mean $\mu_z=[2,3]^\intercal$, a standard deviation along the
            horizontal direction $\sigma_{\mathbf{z}x}=1.0$, a standard
            deviation along the vertical direction $\sigma_{\mathbf{z}y}=2.0$,
            and a correlation coefficient between the vertical and horizontal
            directions $\rho_{\mathbf{z}}=0.7$.

            Plot $\mu_z$, the 95\% confidence ellipse for $\mathbf{z}$, and
            verify that approximately 95\% of the samples lie inside the 95\%
            confidence ellipse.

            You may want to complete the script
            \href{http://github.com/joacorapela/gcnu\_bridging2023}{doExSubmarine\_a.py}
            to address this item.

        \item Select a submarine location $\mathbf{z}$ from those generated in
            the previous item. Sample $N=5$ sonar measurements, assuming the
            submarine is at location $\mathbf{z}$ (i.e., sample from
            $p(\mathbf{y}|\mathbf{z})$ to obtain
            $\mathbf{y}_1,\ldots,\mathbf{y}_N$). Use a standard deviation of
            1.0 for the measurement noise along the horizontal and vertical
            directions, and assume that this noise is uncorrelated along these
            directions.

            Plot $\mathbf{z}$, the 95\% confidence ellipse for the sonar
            measurements given that the submarine located at $\mathbf{z}$,
            and verify that approximately 95\% of the samples lie inside the
            95\% confidence band.

            You may want to complete the script
            \href{http://github.com/joacorapela/gcnu\_bridging2023}{doExSubmarine\_b.py}
            to address this item.

        \item derive a mathematical expression for the posterior of the
            submarine location, given sonar measurements; i.e.,
            $p(\mathbf{z}|\mathbf{y}_1,\ldots,\mathbf{y}_N)$.

            \textit{Hints}: 

            \begin{itemize}

                \item The posterior of the submarine location given sonar
                    measurements is proportional to the joint distribution of
                    the submaring location and sonar measurments; i.e.,
                    $p(\mathbf{z}|\mathbf{y}_1,\ldots\mathbf{y}_N)=\frac{p(\mathbf{y}_1,\ldots,\mathbf{y}_N,\mathbf{z})}{p(\mathbf{y}_1,\ldots,\mathbf{y}_N)}=K\,p(\mathbf{y}_1,\ldots,\mathbf{y}_N,\mathbf{z})$,
                    where $K$ is a value that does not depend on $\mathbf{z}$. Thus, to
                    obtain the posterior we can only keep the terms of the
                    joint that depend on $\mathbf{z}$ and normalize the
                    resulting expression to integrate to one.

                \item The joint is the product of the likelihood and the prior;
                    i.e., 
                    $p(\mathbf{y}_1,\ldots\mathbf{y}_N,\mathbf{z})=p(\mathbf{y}_1,\ldots,\mathbf{y}_N|\mathbf{z})p(\mathbf{z})$.
                    Thus, to only keep the terms of the joint that depend
                    on $\mathbf{z}$, we should only keep the term of the likelihood
                    that depend on $\mathbf{z}$ and combine the result
                    with the prior.

                \item As shown in Claim.~\ref{eq:likelihoodForZ}, the terms of the
                    likelihood that depend on $\mathbf{z}$ are proportional to
                    a Gaussian distribution with mean $\mathbf{z}$ and
                    covariance $\frac{1}{N}\Sigma$; i.e., 
                    $p(\mathbf{y}_1,\ldots,\mathbf{y}_N|\mathbf{z})=K\,\mathcal{N}(\bar{\mathbf{y}}|\mathbf{z},\frac{1}{N}\Sigma_y)$,
                    where $K$ is a value that does not depend on $\mathbf{z}$.

                \item From the previous arguments, to obtain the posterior of
                    $\mathbf{z}$ we can multiply
                    $\mathcal{N}(\bar{\mathbf{y}}|\mathbf{z},\frac{1}{N}\Sigma_y)$
                    with the prior
                    $p(\mathbf{z})=\mathcal{N}(\mathbf{z}|\mu_z,\Sigma_z)$ and
                    normalize the result. To do this we can use the expression
                    for the posterior of the linear Gaussian model described in
                    class.

            \end{itemize}

        \item plot the sample mean of the measurements, the mean of the posterior, its 95\% confidence ellipse, and
            check if the true submarine location, i.e., $\mathbf{z}$, lies
            within this ellipse.

            You may want to complete the script
            \href{http://github.com/joacorapela/gcnu\_bridging2023}{doExSubmarine\_d.py}
            to address this item.

        \item repeat (b) and (d) with $N\in\{3,50,200,500\}$ sonar measurements. How do the
            posterior estimates change with respect as $N$ increases?

        \item write the expressions of the posterior mean and covariances to
            see that:

            \begin{itemize}

                \item as the number of measurements increase, the relative
                    contribution of the prior to estimates of the posterior
                    mean and covariance decreases,

                \item in the limit when the number of measurements approaches
                    infinity, the posterior covariance approaches zero and the
                    posterior mean approaches the measurements sample mean.
                    That is, for an infinite number of measurments, the
                    posterior estimate becomes deterministic and the
                    contribution of the prior to this estimate dissappears.

            \end{itemize}

            Can you see these points in the previous simulations?

    \end{enumerate}

    \begin{comment}
    \section*{Radiation source}

    In a planar 2D surface 5.000 meters below the earth surface there is a source
    of radiation that we aim to discover. We can measure this source with $N=100$
    sensors located at distant points on the 3D earth surface. These measurements
    are noisy because our sensors are noisy. Geophysicist informed us that that the
    distribution of a set of surface measurements $\mathbf{Y}$ given a 2D radiation
    source $\mathbf{X}$ is Gaussian with mean $\mathbf{\mu}=A \mathbf{X} +
    \mathbf{b}$ and covariance $Q$.


    \begin{align*}
        A &= \left[\begin{array}{cc}
                       0.7 & 8.1\\
                       1.05 & 2.5\\
                       0.45 & 2.3
                   \end{array}
             \right]\\
        \mathbf{b} &= \left[\begin{array}{c}
                                0.0\\
                                0.0\\
                                0.0
                            \end{array}
                      \right]\\
        Q &= \left[\begin{array}{cc}
                       0.7 & 0.1\\
                       0.05 & 0.5
                   \end{array}
             \right]
    \end{align*}


    A priory, our guess is that the source of radiation is at location $\mu=(10,
    -3)$ meters, with respect of the center of our parametrization of the deep
    planar surface.  We are equally uncertain about the location of the source of
    radiation in the vertical and horizontal directions. The variance of our
    uncertainty in either direction is 2~$\text{meter}^2$. We know that the source
    of radiation is in a rift, so we assume a correlation coefficient of 0.7 for
    the horizontal and vertical source locations.

    \begin{enumerate}[(a)]

        \item sample $N=100$ instances of sources of radiation and surface
            measurements ($\mathbf{x}_n, \mathbf{y}_n; n=1,\ldots,N$),

        \item plot the mean of the sources of radiation and its 95\% confidence
            ellipse. Verify that 95\% of the sampled sources lie within the 95\%
            confidence ellipse,

        \item for each sampled surface measurement, $\mathbf{y}_n$, infer the mean,
            $\mu_{x|y;n}$, and covariance, $\Sigma_{x|y;n}$, of the the posterior
            probability of the source given the observation,

        \item plot the sampled sources of radiation $\mathbf{x}_{n}$, the estimated
            means $\mu_{x|y;n}$ and $95\%$ confidence ellipses, of the posterior
            distribution of the source of radiation given the surface
            measurements,

    \item count how many times a sampled source of radiation, $\mathbf{x}_n$,
        lies within the 95\% confidence ellipse of mean of the posterior
        probability of the source given a measurement.

    \end{enumerate}
    \end{comment}

\end{document}
